Schrödinger's Equation in a Three-Dimensional Harmonic Oscillator Well \\ \hrulefill\small{ FYS3150: Computational Physics }\hrulefill

### 2018-09-13 Project2/article/article.tex
typos

We now introduce the relative coordinate $\vb{r} = \vb{r}_1 - \vb{r}_2$ and the center of mass coordinate $R=\frac{1}{2}(\vb{r}_1+\vb{r}_2)$ as well as assuming separation of variables $u(r,R)=\psi(r)\phi(R)$. When the energy is given as a sum of the relative energy and center of mass energy, $E^{(2)}=E_r+E_R$ we get
\begin{equation}
\label{eq:radialschr2}
\left(-\frac{\hbar^2}{m}\frac{d^2}{dr^2}+\frac{1}{4}kr^2+\frac{\beta e^2}{r} \right)\psi(r)=E_r\psi(r)
\end{equation}
when also including the repulsive Coulomb interaction, $V(r) = \beta e^2/\abs{\vb{r}_1+\vb{r}_2}= \beta e^2/r$. By once again introducing a dimensionless variable $\rho=r/\alpha$ with $\alpha=\hbar^2/(m\beta e^2)$. By, in addition, including the relative angular frequency $\omega_r^2=(\frac{2\omega}{2\hbar}\alpha^2)^2$ and an energy parameter $\lambda=\frac{m\alpha^2}{\hbar^2}E_r$, we arrive at an equation comparable to \ref{eq:schr3}
\begin{equation}
-\frac{d\psi(\rho)}{d\rho^2}+\left(\omega^2_r\rho^2 + \frac{1}{\rho} \right) \psi(\rho)= \lambda\psi(\rho)
\end{equation}
It is easy to see that the only difference from the situation with two electrons and a single electron is that the potential is changed from $\rho^2$ to $\omega_r^2\rho^2+1/\rho$. In discretization, this is achieved by changing the diagonal of the matrix in \ref{eq:mat1} to $d_i = \frac{2}{hbar^2} \omega_r^2\rho^2+\frac{1}{\rho_i}$.

@@

This proof is important to the algorithm which is described in the next section. As a result of Abel and Ruffini's theorem, regarding the impossibility of a solution to equations degree five and higher, a discretized problema s described herein is unsolvable by conventional, analytical means, and an iterative process is thus necessary.

@@

First, consider a simple two-dimensional, symmetric matrix
\begin{equation}
A = \begin{bmatrix}
a_{11} & a_{12} \\
a_{21} & a_{22}
\end{bmatrix}
\end{equation}
where, because the matrix is symmetric, $a_{12}=a_{21}$. Second, consider an orhtogonal rotation matrix
\begin{equation}
R = \begin{bmatrix}
\cos{\theta} & -\sin{\theta} \\
\sin{\theta} & \cos{\theta}
\end{bmatrix}
\end{equation}
which when multiplied with some vector $\vb{v}$, rotates the vector by an angle $\theta$. Now, we simplify the notation a bit by substituting the trigonometric functions $c=\cos{\theta}$ and $s=\sin{\theta}$. We can now write the similarity transform as
\begin{equation}
R^{-1}AR=\begin{bmatrix}
a_{11}c^2+2a_{12}cs + a_{22}s^2 & (a_{22}-a_{11})cs+a_{12}(c^2-s^2) \\
(a_{22}-a_{11})cs + a_{12}(c^2-s^2) & a_{11}s^2-2a_{12}s+a_{22}c^2
\end{bmatrix}
\end{equation}
Notice that this matrix is symmetric as well. We want to roate in such a manner that the offdiagonal elemtes of the new matrix are zero. Hence we need to solve $(a_{22}-a_{11})cs + a_{12}(c^2-s^2)=0$, which gives
\begin{equation}
\tau = \cot{2\theta} = \frac{a_{11}-a_{22}}{2a_{12}}
\end{equation}

Now we can express all the trigonometric functions in terms of $\tau$. Setting $t = \tan{\theta}$, $c=\cos{\theta}=1/\sqrt{t^2+1}$ and $s=ct$. This yields $\cot{2\theta}=1/2(\cot{\theta}-\tan{\theta}) = 1/2(2^{-1}-t$, all boiling down the quadratic equation
\begin{equation}
t^2-2\tau t-1 =0
\end{equation}
which has two roots. Of these two roots we pickk the smallest in order to ensure maximum  numerical stability. To put it another way, we rotate a tiny angle rather than a large one at each iteration.

@@

The \texttt{jacobiRotation} function works by calculating equation \ref{eq:rotation} straightforward and updating all matrix elements. Lastly the eigenvectors are saved to the matrix R.

>>>>

We now introduce the relative coordinate $\vb{r} = \vb{r}_1 - \vb{r}_2$ and the center of mass coordinate $R=\frac{1}{2}(\vb{r}_1+\vb{r}_2)$ as well as assuming separation of variables $u(r,R)=\psi(r)\phi(R)$. When the energy is given as a sum of the relative energy and center of mass energy, $E^{(2)}=E_r+E_R$ we get
\begin{equation}
\label{eq:radialschr2}
\left(-\frac{\hbar^2}{m}\frac{d^2}{dr^2}+\frac{1}{4}kr^2+\frac{\beta e^2}{r} \right)\psi(r)=E_r\psi(r)
\end{equation}
when also including the repulsive Coulomb interaction, $V(r) = \beta e^2/\abs{\vb{r}_1+\vb{r}_2}= \beta e^2/r$. By once again introducing a dimensionless variable $\rho=r/\alpha$ with $\alpha=\hbar^2/(m\beta e^2)$. By, in addition, including the relative angular frequency $\omega_r^2=(\frac{2\omega}{2\hbar}\alpha^2)^2$ and an energy parameter $\lambda=\frac{m\alpha^2}{\hbar^2}E_r$, we arrive at an equation comparable to \ref{eq:schr2}
\begin{equation}
-\frac{d^2\psi(\rho)}{d\rho^2}+\left(\omega^2_r\rho^2 + \frac{1}{\rho} \right) \psi(\rho)= \lambda\psi(\rho)
\end{equation}
It is easy to see that the only difference from the situation with two electrons and a single electron is that the potential is changed from $\rho^2$ to $\omega_r^2\rho^2+1/\rho$. In discretization, this is achieved by changing the diagonal of the matrix in \ref{eq:mat1} to $d_i = \frac{2}{hbar^2} \omega_r^2\rho^2+\frac{1}{\rho_i}$.

@@

This proof is important to the algorithm which is described in the next section. As a result of Abel and Ruffini's theorem, regarding the impossibility of a solution to equations degree five and higher, a discretized problem as described herein is unsolvable by conventional, analytical means, and an iterative process is thus necessary.

@@

First, consider a simple two-dimensional, symmetric matrix
\begin{equation}
A = \begin{bmatrix}
a_{11} & a_{12} \\
a_{21} & a_{22}
\end{bmatrix}
\end{equation}
where, because the matrix is symmetric, $a_{12}=a_{21}$. Second, consider an orhtogonal rotation matrix
\begin{equation}
R = \begin{bmatrix}
\cos{\theta} & -\sin{\theta} \\
\sin{\theta} & \cos{\theta}
\end{bmatrix}
\end{equation}
which when multiplied with some vector $\vb{v}$, rotates the vector by an angle $\theta$. Now, we simplify the notation a bit by substituting the trigonometric functions $c=\cos{\theta}$ and $s=\sin{\theta}$. We can now write the similarity transform as
\begin{equation}
R^{-1}AR=\begin{bmatrix}
a_{11}c^2+2a_{12}cs + a_{22}s^2 & (a_{22}-a_{11})cs+a_{12}(c^2-s^2) \\
(a_{22}-a_{11})cs + a_{12}(c^2-s^2) & a_{11}s^2-2a_{12}s+a_{22}c^2
\end{bmatrix}
\end{equation}
Notice that this matrix is symmetric as well. We want to roate in such a manner that the off-diagonal elements of the new matrix are zero. Hence we need to solve $(a_{22}-a_{11})cs + a_{12}(c^2-s^2)=0$, which gives
\begin{equation}
\tau = \cot{2\theta} = \frac{a_{11}-a_{22}}{2a_{12}}
\end{equation}

Now we can express all the trigonometric functions in terms of $\tau$. Setting $t = \tan{\theta}$, $c=\cos{\theta}=1/\sqrt{t^2+1}$ and $s=ct$. This yields $\cot{2\theta}=1/2(\cot{\theta}-\tan{\theta}) = 1/2(2^{-1}-t$, all boiling down the quadratic equation
\begin{equation}
t^2-2\tau t-1 =0
\end{equation}
which has two roots. Of these two roots we pick the smallest in order to ensure maximum  numerical stability. To put it another way, we rotate a tiny angle rather than a large one at each iteration.

@@

The \texttt{jacobiRotation} function works by calculating equations \ref{eq:rotation} straightforward and updating all matrix elements. Lastly the eigenvectors are saved to the matrix R.
